---
title: "Feature Engineering — Spotify Tracks Dataset"
subtitle: "MSDS 16:954:567 – Statistical Models and Computing"
author: "Group 16 - Rohan Pant (rp1610) & Pranav Immaneni (pi83)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5,
  dpi        = 72
)
```

## Overview

With the data cleaned and explored, this notebook prepares the final modeling dataset. The steps here are motivated directly by the EDA findings:

- **Duration outliers** were flagged in Section 9 of the EDA — we remove them now.
- **`time_signature`** and **`key`** showed near-zero variance and no relationship with popularity — we drop them.
- **Numeric features** are standardized so that regression coefficients are comparable across features with very different scales (e.g. loudness in dB vs. tempo in BPM).

The output of this notebook — `data/spotify_model.rds` — is the direct input to the PCA and regression scripts.

---

## 1. Load Libraries

```{r libraries}
library(tidyverse)
```

---

## 2. Load Data

```{r load-data}
df <- readRDS("../data/spotify_cleaned.rds")
cat("Loaded:", nrow(df), "rows x", ncol(df), "columns\n")
```

97,696 rows × 19 columns loaded cleanly from the .rds file. Factor levels intact, ready to process.

---

## 3. Remove Duration Outliers

Tracks exceeding 20 minutes (1,200,000 ms) are almost certainly misclassified content — podcasts, DJ sets, live concert recordings — that do not belong in a songs dataset. Including them would give those observations disproportionate leverage in regression.

```{r remove-duration-outliers}
out_dir <- "../outputs/eda"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

save_plot <- function(plot, filename, width = 7, height = 5, dpi = 72) {
  path <- file.path(out_dir, filename)
  ggsave(path, plot = plot, width = width, height = height, dpi = dpi)
  cat("Saved:", path, "\n")
}

# Spotify color palette
spotify_green <- "#1DB954"
spotify_black <- "#121212"
spotify_gray  <- "#535353"
spotify_light <- "#F4F4F4"

base_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14,
                                    margin = margin(b = 8), color = spotify_black),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = spotify_light),
    panel.grid.minor = element_blank(),
    axis.title       = element_text(size = 11, color = spotify_gray),
    axis.text        = element_text(color = spotify_gray),
    strip.text       = element_text(face = "bold", color = spotify_black),
    legend.text      = element_text(color = spotify_gray),
    legend.title     = element_text(color = spotify_black)
  )

n_before <- nrow(df)
df <- df |> filter(duration_ms <= 1200000)
n_removed <- n_before - nrow(df)

cat("Removing", n_removed, "tracks with duration > 20 min...\n")
cat("Remaining rows:", nrow(df), "\n")
```

```{r plot-duration-clean}
mean_dur_s <- mean(df$duration_ms) / 1000

p8 <- ggplot(df, aes(x = duration_ms / 1000)) +
  geom_histogram(
    fill      = spotify_green,
    color     = "white",
    linewidth = 0.2,
    bins      = 60
  ) +
  geom_vline(xintercept = mean_dur_s, color = spotify_black,
             linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = mean_dur_s + 5, y = Inf, vjust = 1.5,
           label = paste0("Mean = ", round(mean_dur_s, 0), "s"),
           color = spotify_black, hjust = 0, size = 3.5) +
  scale_x_continuous(labels = scales::label_number(suffix = "s")) +
  labs(
    title = "Track Duration After Outlier Removal",
    x     = "Duration (seconds)",
    y     = "Count"
  ) +
  base_theme

p8
save_plot(p8, "08_duration_clean.png")
```

---

## 4. Drop Low-Utility Columns

Two columns are dropped based on EDA findings:

- **`time_signature`**: >90% of tracks are in 4/4 time — near-zero variance, no discriminating power.
- **`key`**: 12 levels, no meaningful or consistent relationship with popularity observed in EDA.

```{r drop-columns}
df <- df |> select(-any_of(c("time_signature", "key")))

cat("Remaining columns after dropping:\n")
cat(paste(names(df), collapse = ", "), "\n")
cat("Dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
```

Two columns are removed. *time_signature* exhibited near-zero variance — over 90% of tracks are in 4/4 time, leaving insufficient variation for any model to learn from. *key* had 12 levels but showed no discernible relationship with popularity in EDA. The dataset reduces from 19 to 17 columns.

---

## 5. Separate Metadata and Modeling Columns

Before standardization, we explicitly separate columns into three groups — metadata, factor predictors, and numeric columns. This prevents accidental scaling of categorical variables or the target variable, and makes the modeling pipeline explicit and auditable.

*Metadata columns are kept in the dataset for reference but will not enter any model.*

```{r define-column-groups}
metadata_cols <- c("artists", "album_name", "track_name")
factor_cols   <- c("explicit", "mode", "track_genre")
numeric_cols  <- c(
  "popularity", "danceability", "energy", "loudness", "tempo",
  "acousticness", "speechiness", "instrumentalness", "liveness",
  "valence", "duration_ms"
)

# Summary table
tibble(
  group   = c("Metadata (not modeled)", "Factor predictors", "Numeric columns"),
  columns = c(
    paste(metadata_cols, collapse = ", "),
    paste(factor_cols,   collapse = ", "),
    paste(numeric_cols,  collapse = ", ")
  ),
  n = c(length(metadata_cols), length(factor_cols), length(numeric_cols))
)
```

The dataset is cleanly partitioned into three groups: *3 metadata columns* that won't enter any model, *3 factor predictors* (explicit, mode, track_genre), and *11 numeric columns* including the target (popularity) and control variable (duration_ms). The 9 audio features that will be standardized and fed into PCA are all present.

---

## 6. Standardize Numeric Features

PCA is sensitive to scale. A variable like tempo with values around 120 BPM would dominate a variable like danceability measured between 0 and 1 — not because it's more important, but simply because its numbers are larger. Standardizing to mean = 0, SD = 1 puts all nine features on equal footing before PCA runs.

Note: *popularity and duration_ms are deliberately excluded from standardization*. popularity is our target variable and needs to stay on its original 1–100 scale for interpretability. duration_ms is a control variable we'll handle separately.

```{r standardize}
audio_features <- c(
  "danceability", "energy", "loudness", "tempo",
  "acousticness", "speechiness", "instrumentalness",
  "liveness", "valence"
)

# Compute before-scaling stats for the verification table
before_stats <- df |>
  select(all_of(audio_features)) |>
  summarise(across(everything(), list(mean = mean, sd = sd))) |>
  pivot_longer(everything(),
               names_to  = c("feature", ".value"),
               names_sep = "_(?=[^_]+$)") |>
  rename(mean_before = mean, sd_before = sd)

# Apply scaling and save parameters for potential inverse-transform later
scaling_params <- lapply(audio_features, function(feat) {
  scaled <- scale(df[[feat]])
  list(
    center = attr(scaled, "scaled:center"),
    scale  = attr(scaled, "scaled:scale")
  )
})
names(scaling_params) <- audio_features

saveRDS(scaling_params, "../data/scaling_params.rds")
cat("Scaling parameters saved to ../data/scaling_params.rds\n")

# Apply standardization to the dataframe
df <- df |>
  mutate(across(all_of(audio_features), ~ as.numeric(scale(.))))

# Compute after-scaling stats
after_stats <- df |>
  select(all_of(audio_features)) |>
  summarise(across(everything(), list(mean = mean, sd = sd))) |>
  pivot_longer(everything(),
               names_to  = c("feature", ".value"),
               names_sep = "_(?=[^_]+$)") |>
  rename(mean_after = mean, sd_after = sd)

# Combined verification table
before_stats |>
  left_join(after_stats, by = "feature") |>
  mutate(across(where(is.numeric), ~ round(., 4)))
```

Standardization confirmed across all 9 features — every mean_after is 0 and every sd_after is 1. The before columns show just how different the original scales were: tempo had a mean of 122.7 with SD of 30, while speechiness had a mean of 0.086 with SD of 0.11. Without standardization, PCA would have been completely dominated by tempo and loudness simply due to their larger numeric ranges.

---

## 7. Verify Standardization

All nine features should now be centered at 0 with unit variance.

```{r verify-standardization}
df_std_long <- df |>
  select(all_of(audio_features)) |>
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  mutate(feature = str_to_title(feature))

p9 <- ggplot(df_std_long, aes(x = value)) +
  geom_histogram(
    fill      = spotify_green,
    color     = "white",
    linewidth = 0.2,
    bins      = 40
  ) +
  geom_vline(xintercept = 0, color = spotify_black,
             linetype = "dashed", linewidth = 0.6) +
  facet_wrap(~feature, scales = "free_y", ncol = 3) +
  labs(
    title = "Standardized Audio Features (mean = 0, SD = 1)",
    x     = "Standardized Value",
    y     = "Count"
  ) +
  base_theme +
  theme(axis.text.x = element_text(size = 8))

p9
save_plot(p9, "09_standardized_features.png")
```

All 9 features are now centered at 0, confirmed by the dashed vertical line sitting at the peak of each distribution. The standardization worked correctly across the board.

---

## 8. Final Modeling Dataset

We combine all three column groups — metadata, standardized audio features, and factor predictors — into a single modeling-ready dataset and save it. Every downstream notebook (PCA, regression, bootstrap) loads from this file. One clean source of truth for the entire modeling phase.

```{r final-dataset}
df_model <- df |>
  select(
    all_of(metadata_cols),   # artists, album_name, track_name
    popularity,              # target variable
    all_of(audio_features),  # 9 standardized audio features
    duration_ms,             # control variable (unstandardized)
    all_of(factor_cols)      # explicit, mode, track_genre
  )

cat("Final modeling dataset dimensions:", nrow(df_model), "rows x",
    ncol(df_model), "columns\n\n")

cat("Column names:\n")
cat(paste(names(df_model), collapse = ", "), "\n\n")

head(df_model, 5)
```

```{r save-model-dataset}
saveRDS(df_model, "../data/spotify_model.rds")
write_csv(df_model, "../data/spotify_model.csv")

cat("Saved: ../data/spotify_model.rds\n")
cat("Saved: ../data/spotify_model.csv\n")
```

The final modeling dataset is structured exactly as intended. Metadata columns are present but won't enter any model. popularity is on its original 1–100 scale. Audio features are standardized — visible in the negative values for Ben Woodward's "Ghost (Acoustic)", which scores well below average on energy and loudness, consistent with an acoustic track. Factor columns are intact.

---

*Next: `04_pca.Rmd` — Principal Component Analysis on the standardized audio features.*
